A user-extensible and adaptable parser architecture

### 2008-08-15 doc/sgai-2008.tex
Some suggestions from Carl; more to follow.

\abstract{%
    Some parsers need to be very precise and strict in what they parse, yet
    must allow users to easily adapt or extend the parser to parse new
    inputs, without requiring that the user have an in-depth knowledge and
    understanding of the parser's internal workings.  This paper presents a
    novel parsing architecture which aims to make the process of parsing
    new inputs as simple as possible, enabling users to trivially add new
    rules (to parse variants of existing inputs) and relatively easily add
    new actions (to process a previously unknown class of input).  The
    architecture scales linearly as the number of rules and/or size of
    input increases, making it suitable for parsing large corpora or months
    of accumulated data.
}

@@

XXX DISCUSS COMPLEXITY\@?

@@

It was hoped to reuse an existing parser rather than starting from scratch:
several parsers were considered and rejected because their parsing coverage
was too low, their parsing too inexact, and/or they did not extract
sufficient data.  The effort required to adapt and improve an existing
parser was judged to be greater than the effort to write one, as the
techniques used by the existing parsers severely limited their potential:
some ignored the majority of log lines, parsing specific log lines with
great accuracy, but without any provision for readily parsing new or
similar log lines; others sloppily parsed the majority of log lines, but
were incapable of distinguishing between log lines of the same type, e.g.\
rejections, and extracted only a small portion of the data required.  The
only prior published work on the subject of parsing Postfix log files
is~\cite{log-mail-analyser}, which aims to present correlated log data in a
form suitable for a systems administrator to search using the myriad of
standard Unix text processing utilities already available.  A full state of
the art review is outside the scope of this paper but will be included in
the thesis resulting from this work.

The solution developed is conceptually simple: provide a few generic
functions (\textit{actions\/}), each capable of dealing with an entire
category of log lines (e.g.\ rejections), accompanied by a multitude of
precise patterns (\textit{rules\/}), each of which matches all log lines of
a specific type, and only that type (e.g.\ rejection by a specific DNSBL).
XXX IMPROVE:\@ It is an accepted standard to separate the parsing procedure
from the declarative grammar it operates with.  Part of the novelty here is
that the grammars (rules and actions) are themselves procedures.  This
architecture is ideally suited to parsing inputs where the input is not
fully understood or does not conform to a fixed grammar: the architecture
warns about unparsed inputs and other errors, but continues on parsing as
best it can, allowing the developer of a new parser to decide which
deficiencies are most important and require attention first, rather than
being forced to fix the first error which arises.

@@

Each action contains the code required to deal with a particular type of
log lines, e.g.\ rejections.  The actions are parser-specific: each parser
author will need to write the actions required from scratch, unless
extending an existing parser.  In the Postfix log parser developed for this
project there are \numberOFactions{} actions and \numberOFrules{} rules,
with a very skewed distribution of rules to actions as shown in
\graphref{Distribution of rules per action}.  The distribution is skewed
for the same reason the architecture is so successful in enabling users to
parse new log lines: most new log lines can be parsed by a combination of a
new rule and an existing action.  It is expected that parsers utilising
this architecture will have a high ratio of rules to actions, hopefully
resulting in simpler rules and clearer distinctions between log lines.
\showgraph{build/graph-action-distribution}{Distribution of rules per
action}{Distribution of rules per action} Unsurprisingly the action with
the highest number of associated rules is \texttt{REJECTION}, the action
which handles Postfix rejecting a mail; it is followed by
\texttt{SAVE\_BY\_QUEUEID}, the action responsible for handling informative
inputs: these are log lines that supplement the data gathered from other
log lines (such as rejections), but that do not by themselves represent
interesting events.  The third most common action is, perhaps surprisingly,
\texttt{IGNORE}: the \texttt{IGNORE} action does nothing when executed,
allowing uninteresting log lines to be parsed without causing any effects
(neither \texttt{REJECTION} nor \texttt{IGNORE} imply that the input is
ungrammatical or unparsed).  The framework warns about every unparsed log
line, to alert the user that they need to alter or extend their ruleset;
the Postfix log parser successfully parses every input line in the
\numberOFlogFILESall{} log files it is currently being tested with.  The
remaining actions have only one or two associated rules: in some cases this
is because there will only ever be one type of log line for that action,
e.g.\ all log lines showing that a remote client has connected are matched
by a single rule and handled by the \texttt{CONNECT} action; in other cases
the action is required to address a deficiency in the log files, or a
complication which arises during parsing.  Using the \texttt{CONNECT}
action as an example, it creates a new data structure for each client
connecting, and saves the data extracted by the rule.  If the data
structure already exists it is treated as a symptom of a bug, and the
action issues a warning containing the full contents of the data structure,
plus the log line which has just been parsed.

@@

Warnings are issued for any log lines which are not parsed; no
warnings are issued for unparsed log lines while parsing the
\numberOFlogFILES{} test log files, so it can be safely concluded that
there are zero false negatives.  False positives are harder to quantify:
manually verifying that the correct regex parsed each of the
\numberOFlogLINES{} log lines is infeasible, making it impossible to
quantify the false positive rate; however a random sample of 6039 log lines
(0.00994\% of the total) was parsed and the results manually verified by
inspection to ensure that the correct regex parsed each log line.

Coverage of mails is much more difficult to determine accurately than
coverage of log lines.  The implementation can dump its state tables in a
human readable form; examining these tables with reference to the log files
is the best way to detect mails which were not handled properly.  The
implementation issues warnings when it detects any errors, some of which
may alert the user to a problem.  There should be few or no warnings when
parsing, and when finished parsing the state table should only contain
entries for mails which had yet to be delivered when the log files ended,
or were accepted before the log files began.  A sample of 6000 log lines
was parsed with all debugging options enabled, resulting in 167,448 lines
of output.  All 167,448 lines were examined in conjunction with the input
log file and a dump of the resulting database, verifying that for each of
the input lines the implementation used the correct rule and executed the
correct action, which in turn produced the correct result and inserted the
correct data in the database.  The log segment produced 4 warnings (about
mails which had started before the log file segment), 10 mails remaining in
the state tables, and 1625 connections correctly entered in the database.
No error messages were produced, therefore there were no false negatives.
Given the evidence detailed above, the authors are confident that the false
positive rate when reconstructing a mail is zero.

>>>>

\abstract{%
    Some parsers need to be very precise and strict in what they parse, yet
    must allow users to easily adapt or extend the parser to parse new
    inputs, without requiring that the user have an in-depth knowledge and
    understanding of the parser's internal workings.  This paper presents a
    novel parsing architecture designed for Postfix log files which aims to
    make the process of parsing new inputs as simple as possible, enabling
    users to trivially add new rules (to parse variants of existing inputs)
    and relatively easily add new actions (to process a previously unknown
    class of input).  The architecture scales linearly as the number of
    rules and/or size of input increases, making it suitable for parsing
    large corpora or months of accumulated data.
}

@@

%XXX DISCUSS COMPLEXITY\@?

@@

It was hoped to reuse an existing parser rather than starting from scratch:
several parsers were considered and rejected because their parsing coverage
was too low, their parsing too inexact, and/or they did not extract
sufficient data.  The effort required to adapt and improve an existing
parser was judged to be greater than the effort to write one, as the
techniques used by the existing parsers severely limited their potential:
some ignored the majority of log lines, parsing specific log lines with
great accuracy, but without any provision for readily parsing new or
similar log lines; others sloppily parsed the majority of log lines, but
were incapable of distinguishing between log lines of the same type, e.g.\
rejections, and extracted only a small portion of the data required.  The
only prior published work on the subject of parsing Postfix log files that
we are aware of is~\cite{log-mail-analyser}, which aims to present
correlated log data in a form suitable for a systems administrator to
search using the myriad of standard Unix text processing utilities already
available.  A full state of the art review is outside the scope of this
paper but will be included in the thesis resulting from this work.

The solution developed is conceptually simple: provide a few generic
functions (\textit{actions\/}), each capable of dealing with an entire
category of log lines (e.g.\ rejections), accompanied by a multitude of
precise patterns (\textit{rules\/}), each of which matches all log lines of
a specific type, and only that type (e.g.\ rejection by a specific DNSBL).
XXX IMPROVE:\@ It is an accepted standard to separate the parsing procedure
from the declarative grammar it operates with.  Part of the novelty here is
in the way that the grammars (rules and actions) are themselves procedures.
The relationship to another procedural framework (ATN) is discussed below.
This architecture is ideally suited to parsing inputs where the input is
not fully understood or does not conform to a fixed grammar: the
architecture warns about unparsed inputs and other errors, but continues on
parsing as best it can, allowing the developer of a new parser to decide
which deficiencies are most important and require attention first, rather
than being forced to fix the first error which arises.

@@

XXX RENAME ``REJECTION'' TO ``MAIL\_REJECTED'' AND ``IGNORE'' TO
``UNINTERESTING\_LOG\_LINE'' OR SOMETHING ELSE THAT MAKES IT BRUTALLY
OBVIOUS THE PARSER IS SUCCESSFULLY DEALING WITH THE LOG LINES\@.  I SHOULD
PROBABLY RENAME ``SAVE\_BY\_QUEUEID'' TOO\@.

Each action contains the code required to deal with a particular type of
log lines, e.g.\ rejections.  The actions are parser-specific: each parser
author will need to write the actions required from scratch, unless
extending an existing parser.  In the Postfix log parser developed for this
project there are \numberOFactions{} actions and \numberOFrules{} rules,
with a very skewed distribution of rules to actions as shown in
\graphref{Distribution of rules per action}.  The distribution is skewed
for the same reason the architecture is so successful in enabling users to
parse new log lines: most new log lines can be parsed by a combination of a
new rule and an existing action.  It is expected that parsers utilising
this architecture will have a high ratio of rules to actions, hopefully
resulting in simpler rules and clearer distinctions between log lines.
\showgraph{build/graph-action-distribution}{Distribution of rules per
action}{Distribution of rules per action} Unsurprisingly the action with
the highest number of associated rules is \texttt{REJECTION}, the action
which handles Postfix rejecting a mail; it is followed by
\texttt{SAVE\_BY\_QUEUEID}, the action responsible for handling informative
inputs: these are log lines that supplement the data gathered from other
log lines (such as rejections), but that do not by themselves represent
interesting events.  The third most common action is, perhaps surprisingly,
\texttt{IGNORE}: the \texttt{IGNORE} action does nothing when executed,
allowing uninteresting log lines to be parsed without causing any effects
(neither \texttt{REJECTION} nor \texttt{IGNORE} imply that the input is
ungrammatical or unparsed).  The framework warns about every unparsed log
line, to alert the user that they need to alter or extend their ruleset;
the Postfix log parser successfully parses every input line in the
\numberOFlogFILESall{} log files it is currently being tested with.  The
remaining actions have only one or two associated rules: in some cases this
is because there will only ever be one type of log line for that action,
e.g.\ all log lines showing that a remote client has connected are matched
by a single rule and handled by the \texttt{CONNECT} action; in other cases
the action is required to address a deficiency in the log files, or a
complication which arises during parsing.  Taking the \texttt{CONNECT}
action as an example, it creates a new data structure for each client
connecting, and saves the data extracted by the rule (in memory,
subsequently to be entered into the database).  If the data structure
already exists it is treated as a symptom of a bug, and the action issues a
warning containing the full contents of the data structure, plus the log
line which has just been parsed.

@@

Warnings are issued for any log lines which are not parsed; no warnings are
issued for unparsed log lines while parsing the \numberOFlogFILES{} test
log files, so it can be safely concluded that there are zero false
negatives.  False positives are harder to quantify: manually verifying that
the correct regex parsed each of the \numberOFlogLINES{} log lines is
infeasible, making it impossible to quantify the false positive rate.  A
random sample of 6039 log lines (0.00994\% of the total) was parsed and the
results manually verified by inspection to ensure that the correct regex
parsed each log line.  The sample contained zero false positives, and this
check has been automated to ensure continued accuracy.

Coverage of mails is much more difficult to determine accurately than
coverage of log lines.  The implementation can dump its state tables in a
human readable form; examining these tables with reference to the log files
is the best way to detect mails which were not handled properly.  The
implementation issues warnings when it detects any errors, some of which
may alert the user to a problem.  There should be few or no warnings when
parsing, and when finished parsing the state table should only contain
entries for mails which had yet to be delivered when the log files ended,
or were accepted before the log files began.  A second sample of 6000 log
lines was parsed with all debugging options enabled, resulting in 167,448
lines of output.  All 167,448 lines were examined in conjunction with the
input log file and a dump of the resulting database, verifying that for
each of the input lines the implementation used the correct rule and
executed the correct action, which in turn produced the correct result and
inserted the correct data in the database.  The log segment produced 4
warnings (about mails which had started before the log file segment), 10
mails remaining in the state tables, and 1625 connections correctly entered
in the database.  No error messages were produced, therefore there were no
false negatives.  Given the evidence detailed above, the authors are
confident that the false positive rate when reconstructing a mail is zero.
